### 2025-05-08 thesis/thesis.tex
Comparison

% PICARD etc.

@@

>>>>

A major challenge in NL2SQL research is making sure that model generated queries are not just semantically accurate
but also syntactically valid queries and thus executable. To address this issue \citeauthor*{PICARD} released \textsc{Picard}
(Parsing Incrementally for Constrained Auto-Regressive Decoding) in \citeyear{PICARD}, a constrained decoding mechanism
for language models which utilizes the SQL grammar and constrained decoding mechanisms to incrementally parse the generate SQL,
rejecting invalid tokens based on the grammar. \textsc{Picard} showed to significantly improve the performance of pre-trained
language models (like T5 or BERT) when it comes to NL2SQL tasks, lifting them from mid-level to state-of-the-art solutions
on the \textsc{Spider} benchmark \citep{PICARD}.

\textsc{Picard} operates as a incremental parser during model output decoding of pre-trained language models and continuously
evaluates the probability of each token. Instead of just passing model outputs to a database for execution \textsc{Picard}
incrementally parses and validates the generated SQL, rejecting tokens if neeeded thus significantly improving the valid
output accuracy (sometimes referred to as VA) of language models. This approach is addressing a significant issue associated
with pre-trained language models — while they outperform in natural language understanding and reasoning, they often lack
SQL grammer knowledge and tend to generate queries that are not executable due to ther unconstrained output space \citep{PICARD}.

The above introduces RESDSQL built ontop of \textsc{Picard}'s foundations and used a ranking-enhanced framework for input
encoding. These two approaches represent a unqiue class of approaches that utilize input and output constraining in order
to increase the performance characteristics of pre-trained language models \citep{RESDSQL}.

\subsubsection{Advantages of PLM Approaches}

PLM approaches to NL2SQL tasks have yielded significant performance improvements for the NL2SQL domain and represent a
leap in NLIDB-research. They primed the research field towards using language models which led to a transition towards
large language models in the following years. Namely PLM approaches brought a series of upsides with them:

\begin{enumerate}
    \item \textbf{Compute Efficiency} — PLMs like \textsc{RESDSQL} achieve high accuracy (up to 84.1\% on \textsc{Spider}
        depending on variants) wilst using far fewer parameters than conteporary LLMs, making them significantly more efficient
        and therefore reduce hardware requirements for their deployment \citep{RESDSQL}.
    \item \textbf{Transferability} — Approaches like \textsc{Grappa} and \textsc{StruG} can incorporate domain-specific understanding
        of natural language, table structures and SQL syntax during pre-training which addressed one of the primary issues with neural
        approaches \citep{GRAPPA, STRUG}.
    \item \textbf{Vocabulary} — PLMs offer a larger vocabulary due to the vast amounts of training data available. This enables them to
        handle a wide variety of natural language patterns which addresses the benchmark-overfitting tendency of neural approaches which
        primarily trained on the development sets of contemporary benchmarks.
\end{enumerate}

\subsubsection{Limitations of PLM Approaches}

Although representing the state-of-the-art at the time, PLMs introduce a class of problems which are associated with their
non-NL2SQL associated nature. There have been an array of approaches to mitigate these shortcommings but nonetheless they
must be considered when using a PLM-based approach to NL2SQL:

\begin{enumerate}

    \item \textbf{Fine-tuning Requirements} — Most PLMs require substantial domain-specific, or at least NL2SQL specific, fine-tuning, limiting a 
          straight forward adaptation to new domains or databases. Although being significantly more efficient than LLM-based approaches the potential need 
          for initial fine-tuning represent a significant computational resource burden. Furthermore when not using synthetic data generation (e.g. 
          \textsc{Grappa}) annotated datasets of training data are needed to achieve appropiate performance characteristics \citep{GRAPPA}.
    \item \textbf{Wide Input \& Output Space} — Due to the general nature of PLMs their input and output space is often far larger than needed 
          NL2SQL tasks. ``Large pre-trained language models for textual data have an unconstrained output space; at each decoding step, they can
          produce any of 10,000s of sub-word tokens'' \citep{PICARD}. This applies to both the input and output token space, therefore multiple
          approaches have been researched which focus on constraining these to the subset needed for NL2SQL tasks. Namely GRAPHIX-T5 and \textsc{Picard}
          have proposed potential (and promising) solutions to this issue \citep{GRAPHIX, PICARD}.
    \item \textbf{Limited Schema Awareness} — Due to being general purpose, and non-NL2SQL optimized, PLMs tend to incorporate limited amounts
          of schema awareness when being applied out of the box for NL2SQL tasks. Multiple research efforts focused on improving this situation,
          most notably \textsc{Resdsql} and GRAPHIX-T5 tried to improve the schema linking \& awareness of PLMs \citep{RESDSQL, GRAPHIX},
          nonetheless the non-specialized nature of PLMs prevents NL2SQL being part of the fundamental model architecture.
\end{enumerate}

These characteristics positioned PLMs as powerful but comparatively resource-intensive solutions for NL2SQL (especially in direct comparison with neural 
approaches), ultimately yielding the research domain to transition toward exploring Large Language Model approaches that promise even greater flexibility
in adaptation and potentially superior handling of complex queries through advanced in-context learning approaches.

@@

% https://raw.githubusercontent.com/HKUSTDial/NL2SQL_Handbook/refs/heads/main/assets/river.svg
% https://arxiv.org/abs/2109.05093
